---
title: "Do Harsher DUI Punishments Reduce Recidivism?"
subtitle: "Evidence from a Regression Discontinuity Design"
author: "DTC Group 1"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    width: 1280
    height: 720
echo: false
warning: false
message: false
---

## Research Question & Data

**Do harsher punishments for drunk driving reduce repeat offenses?**

-   At BAC = 0.08, penalties jump sharply in Washington State
-   128,603 traffic stops from 1999-2007
-   Running variable: minimum of two BAC tests (legally binding value)
-   Outcome: whether the driver committed another DUI offense

We use a **Regression Discontinuity Design (RDD)** — comparing drivers just above and just below the 0.08 threshold, who are essentially the same people but received different punishments.

## Why RDD Works Here

-   Drivers can't precisely control their BAC at the cutoff
-   Someone at 0.079 vs 0.081 is the same type of drinker
-   The **only difference** is punishment severity
-   Any jump in recidivism at 0.08 is caused by the punishment

**Key choices:**

-   Local linear model
-   Sample restricted below 0.15 to isolate the 0.08 effect
-   Heteroskedasticity-robust standard errors (binary outcome)

## Assumption Tests

::::: columns
::: {.column width="50%"}
**McCrary Density Test**

No manipulation detected (p = 0.23)

Heaping at exactly 0.08 is rounding, not strategic behavior

*Sensitivity check:* Dropping observations within ±0.001 of cutoff — results hold
:::

::: {.column width="50%"}
**Covariate Balance**

| Variable | p-value |
|----------|---------|
| Male     | 0.63    |
| White    | 0.98    |
| Age      | 0.06    |
| Accident | 0.54    |

No significant jumps — drivers on both sides are comparable
:::
:::::

## RD Plot: Visual Evidence

```{r}
library(tidyverse)
library(rdrobust)
library(rddensity)
library(fixest)

c_cut <- 0.08
dwi_raw <- readr::read_csv("DWI_Data.csv")

dwi <- dwi_raw %>%
  mutate(
    bac_min = pmin(bac1, bac2, na.rm = TRUE),
    treat   = as.integer(bac_min >= c_cut),
    x       = bac_min - c_cut
  ) %>%
  filter(!is.na(bac_min), !is.na(recidivism)) %>%
  filter(bac_min < 0.15)

dwi_plot <- dwi %>% filter(bac_min >= 0.03, bac_min <= 0.13)
dwi_main <- dwi %>% filter(bac_min >= 0.05, bac_min <= 0.11)
dwi_narrow <- dwi %>% filter(bac_min >= 0.06, bac_min <= 0.10)
dwi_wide <- dwi %>% filter(bac_min >= 0.04, bac_min <= 0.12)

bins <- dwi_main %>%
  mutate(bac_bin = round(bac_min, 3)) %>%
  group_by(bac_bin) %>%
  summarise(recidivism = mean(recidivism), .groups = "drop")

check_main <- feols(recidivism ~ treat + x + treat:x, data = dwi_main, vcov = "hetero")

grid <- tibble(bac_min = seq(0.05, 0.11, by = 0.001)) %>%
  mutate(x = bac_min - c_cut, treat = ifelse(bac_min >= c_cut, 1, 0))
grid$yhat <- predict(check_main, newdata = grid)

ggplot(bins, aes(x = bac_bin, y = recidivism)) +
  geom_point(color = "steelblue", size = 2) +
  geom_vline(xintercept = c_cut, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_line(data = grid, aes(x = bac_min, y = yhat), linewidth = 1) +
  labs(x = "Blood Alcohol Content (BAC)", y = "Mean Recidivism Rate",
       title = "Recidivism vs BAC at the 0.08 Legal Threshold") +
  theme_minimal(base_size = 16)
```

## Results: Parametric Estimates

```{r}
check_narrow <- feols(recidivism ~ treat + x + treat:x, data = dwi_narrow, vcov = "hetero")
check_wide <- feols(recidivism ~ treat + x + treat:x, data = dwi_wide, vcov = "hetero")
check_cov <- feols(recidivism ~ treat + x + treat:x + male + white + aged + factor(year), data = dwi_main, vcov = "hetero")

etable(check_main, check_narrow, check_wide, check_cov,
       headers = c("Main", "Narrow", "Wide", "Controls"),
       dict = c(treat = "Above 0.08 Cutoff", x = "BAC (centered)"),
       title = "Effect of Harsher DUI Punishment on Recidivism")
```

Treatment effect is **consistently negative** (-0.014 to -0.016) across all bandwidths. Adding controls barely changes the estimate — confirms RDD validity.

## Key Finding

::::: columns
::: {.column width="50%"}
### Absolute effect

Crossing the 0.08 threshold reduces recidivism by **1.5 to 2 percentage points**

For every 100 drivers punished, \~2 fewer reoffend
:::

::: {.column width="50%"}
### Relative effect

Given a baseline rate of \~11%, this is a **14-18% relative reduction**

Consistent across all specifications and bandwidths
:::
:::::

**Bottom line:** Harsher DUI punishments at the 0.08 BAC threshold meaningfully deter repeat drunk-driving offenses.

## Limitations

-   **Washington State only** — may not generalize to other states
-   **Only 0.08 threshold** — the 0.15 cutoff may show different effects
-   **Cannot isolate punishment components** — fines, license suspension, and jail are bundled
-   **Heaping at cutoff** — addressed by sensitivity check, but warrants caution
-   **Possible undercount** — undetected repeat offenses are not captured

## Conclusion

**Harsher DUI punishments reduce recidivism.**

-   1.5-2 percentage point reduction (14-18% relative)
-   Robust across bandwidths, controls, and functional forms
-   RDD assumptions validated by density test and covariate balance
-   Policy implication: stricter BAC enforcement deters repeat offenses

**This is a credible causal estimate, not an artifact of modeling choices.**
