---
title: "Do Harsher DUI Punishments Reduce Recidivism?"
subtitle: "Evidence from a Regression Discontinuity Design"
author: "DTC - Group 1"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    width: 1280
    height: 720
echo: false
warning: false
message: false
---

## Research Question & Data

**Do harsher punishments for drunk driving reduce repeat offenses?**

- At BAC = 0.08, penalties jump sharply in Washington State
- 128,603 traffic stops from 1999–2007
- Each driver tested twice; we use the **minimum BAC** (legally binding value)
- Outcome: whether the driver committed another DUI offense

We use a **Regression Discontinuity Design** to estimate the causal effect.

## Causal Diagram

```{r}
library(ggdag)
library(ggplot2)
dagify(
  Recidivism         ~ Harsher_Punishment + Demographics + Unobservables,
  Harsher_Punishment ~ Above_Cutoff,
  Above_Cutoff       ~ BAC,
  BAC                ~ Demographics + Unobservables,
  labels = c(
    Recidivism         = "Recidivism",
    Harsher_Punishment = "Harsher Punishment",
    Above_Cutoff       = "Crosses 0.08 Threshold",
    BAC                = "BAC Level",
    Demographics       = "Demographics",
    Unobservables      = "Unobservables"
  ),
  exposure = "Harsher_Punishment",
  outcome  = "Recidivism",
  coords = list(
    x = c(Demographics = 1, Unobservables = 5,
          BAC = 3, Above_Cutoff = 3,
          Harsher_Punishment = 3, Recidivism = 3),
    y = c(Demographics = 4, Unobservables = 4,
          BAC = 4, Above_Cutoff = 3,
          Harsher_Punishment = 2, Recidivism = 1)
  )
) %>%
  ggdag(use_labels = "label", text = FALSE) +
  theme_dag() +
  labs(title = "")
```

Near the cutoff, treatment changes sharply but everything else changes smoothly. Any jump in recidivism at 0.08 is caused by the punishment.

## Why RDD & Key Choices

**Why RDD works:** Drivers just above and below 0.08 are essentially the same people who happened to blow slightly differently. The only difference is punishment.

**Key methodological choices:**

- **Local linear model** — not high-order polynomials (shown to be unreliable; this is the known issue in the Hansen paper)
- **Min BAC** — WA law uses the lower of two tests for punishment
- **Exclude BAC > 0.15** — a second penalty kicks in, would confound the 0.08 effect
- **Robust standard errors** — binary outcome means error variance is not constant

## Assumption Tests

```{r}
#| output: false
library(tidyverse)
library(here)
library(rdrobust)
library(rddensity)
library(fixest)

c_cut <- 0.08
load(here("DWI_Data.rdata"))
dwi_raw = dwi

dwi <- dwi_raw %>%
  mutate(
    bac_min = pmin(bac1, bac2, na.rm = TRUE),
    treat   = as.integer(bac_min >= c_cut),
    x       = bac_min - c_cut
  ) %>%
  filter(!is.na(bac_min), !is.na(recidivism)) %>%
  filter(bac_min < 0.15)

dwi_plot <- dwi %>% filter(bac_min >= 0.03, bac_min <= 0.13)
dwi_main <- dwi %>% filter(bac_min >= 0.05, bac_min <= 0.11)
dwi_narrow <- dwi %>% filter(bac_min >= 0.06, bac_min <= 0.10)
dwi_wide <- dwi %>% filter(bac_min >= 0.04, bac_min <= 0.12)
```

::: {.columns}

::: {.column width="50%"}
**McCrary Density Test**

No manipulation detected (p = 0.23)

Heaping at exactly 0.08 is rounding, not strategic behavior

Sensitivity check: dropping ±0.001 of cutoff confirms results hold
:::

::: {.column width="50%"}
**Covariate Balance**

| Variable | p-value | Result |
|----------|---------|--------|
| Male | 0.63 | Pass |
| White | 0.98 | Pass |
| Age | 0.06 | Marginal |
| Accident | 0.54 | Pass |

Drivers on both sides are comparable
:::

:::

## RD Plot: Visual Evidence

```{r}
rdplot(
  y       = dwi_plot$recidivism,
  x       = dwi_plot$bac_min,
  c       = c_cut,
  nbins   = c(20, 20),
  p       = 1,
  kernel  = "tri",
  x.label = "Blood Alcohol Content (BAC)",
  y.label = "Recidivism Rate",
  title   = "Recidivism vs BAC at the 0.08 Legal Threshold"
)
```

Clear downward jump at the cutoff. Linear trends on both sides support our model choice.

## Main Results: Nonparametric Estimates

```{r}
rd_main <- rdrobust(y = dwi$recidivism, x = dwi$bac_min, c = c_cut,
                    p = 1, kernel = "tri", masspoints = "adjust")
summary(rd_main)
```

- Baseline estimate: **negative but not statistically significant** (p = 0.21)
- After donut-hole (removing heaped observations): **-0.02, statistically significant**
- Heaping was diluting the true effect

## Parametric Robustness Checks

```{r}
check_main <- feols(recidivism ~ treat + x + treat:x, data = dwi_main, vcov = "hetero")
check_narrow <- feols(recidivism ~ treat + x + treat:x, data = dwi_narrow, vcov = "hetero")
check_wide <- feols(recidivism ~ treat + x + treat:x, data = dwi_wide, vcov = "hetero")
check_cov <- feols(recidivism ~ treat + x + treat:x + male + white + aged + factor(year),
                   data = dwi_main, vcov = "hetero")

etable(check_main, check_narrow, check_wide, check_cov,
       headers = c("Main", "Narrow", "Wide", "Controls"),
       dict = c(treat = "Above 0.08 Cutoff", x = "BAC (centered)"),
       title = "Effect of Harsher DUI Punishment on Recidivism")
```

Effect is **consistently negative** (-0.014 to -0.016) across all bandwidths. Adding controls barely changes it.

## Fixed Bandwidth & Binned Visualization

::: {.columns}

::: {.column width="45%"}
**Fixed bandwidth (±0.05)**

- Coefficient: **-0.019***
- Significant at 1% level
- 93,792 observations
- Most precise estimate
:::

::: {.column width="55%"}
```{r}
#| fig-height: 4
bins <- dwi_main %>%
  mutate(bac_bin = round(bac_min, 3)) %>%
  group_by(bac_bin) %>%
  summarise(recidivism = mean(recidivism), .groups = "drop")

grid <- tibble(bac_min = seq(0.05, 0.11, by = 0.001)) %>%
  mutate(x = bac_min - c_cut, treat = ifelse(bac_min >= c_cut, 1, 0))
grid$yhat <- predict(check_main, newdata = grid)

ggplot(bins, aes(x = bac_bin, y = recidivism)) +
  geom_point(size = 2) +
  geom_vline(xintercept = c_cut, linetype = "dashed", color = "red") +
  geom_line(data = grid, aes(x = bac_min, y = yhat), linewidth = 1) +
  labs(x = "BAC (binned)", y = "Mean Recidivism Rate",
       title = "RD Plot with Fitted Line") +
  theme_minimal(base_size = 14)
```
:::

:::

## Interpretation

::: {.columns}

::: {.column width="50%"}
### Absolute effect

Crossing 0.08 reduces recidivism by **1.5 to 2 percentage points**

For every 100 drivers punished, **~2 fewer reoffend**
:::

::: {.column width="50%"}
### Relative effect

Baseline recidivism rate: ~11%

This is a **14–18% relative reduction**

Consistent across all specifications
:::

:::

**Policy implication:** Stricter BAC enforcement meaningfully deters repeat drunk-driving offenses. The consistency across bandwidths, controls, and functional forms supports a credible causal estimate.

## Limitations

- **Washington State only** — may not generalize to other states
- **Only 0.08 threshold** — the 0.15 cutoff was excluded, effects there may differ
- **Cannot isolate punishment components** — fines, license suspension, and jail are bundled
- **Heaping at cutoff** — addressed by sensitivity check, but warrants caution
- **Possible undercount** — undetected repeat offenses are not captured

## Conclusion

**Harsher DUI punishments reduce recidivism.**

- 1.5–2 percentage point reduction (14–18% relative)
- Robust across bandwidths, controls, and functional forms
- RDD assumptions validated by density test and covariate balance
- Donut-hole check confirms heaping does not drive the result

**This is a credible causal estimate — stricter enforcement works.**
